Allosteric activation or inhibition of PI3Kγ mediated through conformational changes in the p110γ helical domain

Version 1

Abstract

Abstract

PI3Kγ is a critical immune signaling enzyme activated downstream of diverse cell surface molecules, including Ras, PKCβ activated by the IgE receptor, and Gβγ subunits released from activated GPCRs. PI3Kγ can form two distinct complexes, with the p110γ catalytic subunit binding to either a p101 or p84 regulatory subunit, with these complexes being differentially activated by upstream stimuli. Here using a combination of Cryo electron microscopy, HDX-MS, and biochemical assays we have identified novel roles of the helical domain of p110γ in regulating lipid kinase activity of distinct PI3Kγ complexes. We defined the molecular basis for how an allosteric inhibitory nanobody potently inhibits kinase activity through rigidifying the helical domain and regulatory motif of the kinase domain. The nanobody did not block either p110γ membrane recruitment or Ras/Gβγ binding, but instead decreased ATP turnover. We also identified that p110γ can be activated by dual PKCβ helical domain phosphorylation leading to partial unfolding of an N-terminal region of the helical domain. PKCβ phosphorylation is selective for p110γ-p84 compared to p110γ-p101, driven by differential dynamics of the helical domain of these different complexes. Nanobody binding prevented PKCβ mediated phosphorylation. Overall, this works shows an unexpected allosteric regulatory role of the helical domain of p110γ that is distinct between p110γ-p84 and p110γ-p101, and reveals how this can be modulated by either phosphorylation or allosteric inhibitory binding partners. This opens possibilities of future allosteric inhibitor development for therapeutic intervention.

Introduction

The class I phosphoinositide 3 kinases (PI3Ks) are master regulators of myriad functions through their generation of the lipid signalling molecule phosphatidylinositol 3,4,5-trisphosphate (PIP3) downstream of cell surface receptors (Burke and Williams, 2015; Rathinaswamy and Burke, 2019; Vanhaesebroeck et al., 2021; Vasan and Cantley, 2022). The class I PI3Ks can be further subdivided into the class IA and class IB subfamilies, with class IB PI3Ks being critical in immune signalling, and are composed of a single p110γ catalytic subunit that can bind to either a p101 or p84 regulatory subunit (Hawkins and Stephens, 2015; Lanahan et al., 2022; Okkenhaug, 2013). The two PI3Kγ complexes (either p110γ-p84 or p110γ-p101) play essential and independent roles in both the adaptative and innate immune systems. PI3Kγ has shown promise as a therapeutic target, primarily as an immunomodulator of the tumor microenvironment leading to enhanced anti-tumor immune responses (De Henau et al., 2016; Kaneda et al., 2016b). Multiple isoform selective small molecule ATP competitive inhibitors of p110γ are in clinical trials for multiple forms of human cancers (Li et al., 2021). However, all inhibitors currently developed towards p110γ act as ATP competitive inhibitors, showing equal potency against both p110γ-p84 or p110γ-p101 complexes.

Detailed experiments on the role of p110γ in mice show that knockout of both p101 and p84 leads to PIP3 levels that are equivalent to knockout of p110γ, showing that all cellular PI3Kγ activity requires the presence of either a p84 or p101 regulatory subunit (Rynkiewicz et al., 2020). The two complexes are differentially activated by membrane localised receptors, including G-protein coupled receptors (Li et al., 2000; Stephens et al., 1997), Ras (Jin et al., 2020; Kurig et al., 2009), toll like receptors (TLRs) (Luo et al., 2018), and the IgE antigen receptor (Laffargue et al., 2002; Walser et al., 2013). This leads to the different complexes driving unique immune responses, with p110γ-p101involved in chemotaxis in neutrophils (Bohnacker et al., 2009; Deladeriere et al., 2015), and p110γ-p84 involved in reactive oxide production. Differential activation of unique PI3Kγ complexes downstream of GPCRs and Ras is caused by the ability of p101 to directly bind to Gβγ subunits downstream of activated GPCRs, with this being lost in p84, making p110γ-p84 activation by Gβγ dependent on Ras mediated membrane recruitment (Rathinaswamy et al., 2023; Kurig et al., 2009; Rynkiewicz et al., 2020). Activation of PI3Kγ downstream of the IgE antigen receptor is driven by calcium mediated activation of protein kinase Cβ, leading to the selective phosphorylation and activation of p110γ at S582 (Walser et al., 2013), with this putatively only occurring in p110γ-p84 and not p110γ-p101. The full molecular mechanisms underlying how phosphorylation of p110γ is selective for different p84 or p101 complexes, and how it activates lipid kinase activity are poorly understood.

Extensive biophysical approaches including cryo electron microscopy (cryo-EM), X-ray crystallography, and hydrogen deuterium exchange mass spectrometry (HDX-MS) have provided extensive insight into the molecular underpinnings of how p110γ associates with both p101 and p84, how they are differentially activated by Ras and GPCR signals, and how they can be activated on lipid membranes (Pacold et al., 2000; Walker et al., 1999; Rathinaswamy et al., 2021c, 2021a; Gangadhara et al., 2019; Vadas et al., 2013; Rathinaswamy et al., 2023). The p110γ catalytic subunit is composed of an adaptor binding domain (ABD), a Ras binding domain (RBD), a C2 domain, a helical domain, and a bi-lobal kinase domain (Rathinaswamy et al., 2021a; Walker et al., 1999). A set of helices positioned C-terminal to the activation loop in the kinase domain play a critical role in regulating activity, with this region referred to as the regulatory motif (Rathinaswamy et al., 2021c). The p110γ isoform is unique in that it is inhibited in the absence of a regulatory subunit, with this driven by an autoinhibitory conformation of the regulatory motif, that is proposed to require membrane association to disrupt (Gangadhara et al., 2019). The regulatory motif is a common site of activating mutations in the other class I PI3K isoforms (Jenkins et al., 2023), with p110γ having rare activating oncogenic mutants in this region (Rathinaswamy et al., 2021c). The p110γ subunit interacts with both p84 and p101 at an interface composed of the C2 domain, and the linkers between the RBD-C2 and C2-helical domains (Rathinaswamy et al., 2023, 2021a). The p110γ-p84 complex forms a more dynamic complex compared to p110γ-p101 (Rathinaswamy et al., 2023; Shymanets et al., 2013), however, no clear unique regulatory role of this difference in dynamics has been identified.

The fundamental roles of p110γ in inflammatory processes has made it a therapeutic target in many pathological conditions, including asthma (Campa et al., 2018), arthritis (Camps et al., 2005), obesity (Becattini et al., 2011; Breasson et al., 2017), and cancer (De Henau et al., 2016; Kaneda et al., 2016a, 2016b). There are significant side effects from non-isoform selective PI3K inhibitors (Bohnacker et al., 2017; Vanhaesebroeck et al., 2021), which has driven the development of highly p110γ selective small molecule inhibitors (Bell et al., 2012; Evans et al., 2016; Gangadhara et al., 2019). However, all p110γ inhibitors will target both p110γ-p101 and p110γ-p84, so there is a potential for the development of allosteric inhibitors outside of the ATP binding site. Initial promise has been reported for the development of class IA p110α oncogene specific allosteric inhibitors. However, further investigation of the molecular mechanisms underlying p110γ regulation will be required for the discovery of regions that can be targeted for allosteric inhibitor development.

Here we report critical roles of the helical domain of p110γ in both activation and inhibition of lipid kinase activity. We characterised an allosteric inhibitory nanobody (NB7) that potently inhibits p110γ activity. Cryo-EM was used to define the inhibitory interface, which is composed of the helical domain, the ABD-RBD linker, and the regulatory motif of the kinase domain of p110γ. The region that the nanobody binds to is in close spatial proximity to a previously identified PKCβ phosphorylation site (S582) in the helical domain, and oncogenic activating mutants in the regulatory motif. We fully characterised the activity and dynamics of stoichiometrically PKCβ phosphorylated p110γ, leading to the discovery of a novel additional phosphorylation site (S594/S595). PKCβ phosphorylation was highly selective for p110γ and p110γ-p84, with limited phosphorylation of p110γ-p101. Hydrogen deuterium exchange mass spectrometry (HDX-MS) analysis showed that phosphorylation of p110γ leads to unfolding of the N-terminal region of the helical domain, and increased kinase activity. The presence of the inhibitory nanobody significantly blocks PKCβ phosphorylation, with phosphorylated p110γ showing no detectable binding to NB7. Overall, this work provides unique insight into the critical role of the helical domain in controlling p110γ activity, and how phosphorylation and binding partners can modify this regulation. It also reveals a unique binding site located at the interface of the helical and kinase domain that can be targeted for future allosteric inhibitor design.

Results

Molecular mechanism of nanobody inhibition of p110γ

We previously identified multiple nanobodies that inhibited the activity of p110γ. One from this group (denoted NB7 throughout the manuscript) potently inhibited the membrane mediated activation of p110γ-p84 by both Ras and Gβγ, with HDX-MS experiments mapping the NB7 binding interface to the RBD, helical and kinase domains (Rathinaswamy et al., 2021b). We originally hypothesized that NB7 worked by sterically inhibiting Ras binding to the RBD domain of p110γ. To further explore the molecular mechanism of inhibition we purified all complexes of p110γ (p110γ apo, p110γ-p84, p110γ-p101) along with the NB7 nanobody. The SDS-PAGE of all proteins utilised in this study are shown in the source data file included in the supplemental information.

To define the mechanism concerning how NB7 inhibits PI3K activity we analyzed how this nanobody inhibited all class IB PI3K complexes (p110γ, p110γ-p84, p110γ-p101) upon activation by lipidated Gβγ subunits. Intriguingly, we found that all three forms of p110γ were potently inhibited by NB7 (Fig. 1A). While the IC50 measured for the three complexes was different, this is likely mainly due to the dramatic difference in protein required to measure lipid kinase activity in vitro (∼300 nM for p110γ apo/p110γ-p84, and ∼10 nM for p110γ-p101, respectively). This suggested that the mechanism of inhibition was not driven by a steric block of Ras association through the RBD, as previously proposed (Rathinaswamy et al., 2021b). We examined the binding of this nanobody to all complexes using biolayer interferometry (BLI). The nanobody bound equivalently tightly to all complexes, with ∼2 nM potency for p110γ, p110γ-p84, and p110γ-p101 (Fig. 1B). We also tested binding of the nanobody to all class IA PI3Ks, and there was no detectable binding to p110α, p110β, and p110δ (Fig. 1C).

The inhibitory nanobody NB7 binds tightly to all p110γ complexes and inhibits kinase activity, but does not prevent membrane binding

A. Cartoon schematic depicting nanobody inhibition of activation by lipidated Gβγ (1.5 µM final concentration). Lipid kinase assays show a potent inhibition of lipid kinase activity with increasing concentrations of NB7 (3-3000 nM) for the different complexes. The protein concentration of p110γ (300 nM), p110γ-p84 (330 nM) and p110γ-p101 (12 nM) was different due to intrinsic differences of each complex to be activated by lipidated Gβγ.

B. Association and dissociation curves for the dose response of His-NB7 binding to p110γ, p110γ-p84 and p110γ-p101 (50 – 1.9 nM) is shown. A cartoon schematic of BLI analysis of the binding of immobilized His-NB7 to p110γ is shown on the left. Dissociation constants (KD) were calculated based on a global fit to a 1:1 model for the top three concentrations and averaged with error shown.

B. Association and dissociation curves for His-NB7 binding to p110γ, p110α-p85α, p110β-p85β, and p110δ-p85β. Experiments were performed in duplicate with a final concentration of 50 nM of each class I PI3K complex.

D. Total Internal Reflection Fluorescence Microscopy (TIRF-M) analysis of the effect of nanobody NB7 on PI3K recruitment to supported lipid bilayers containing H-Ras(GTP) and farnesyl-Gβγ. Y647-p84/p110γ displays rapid equilibration kinetics and is insensitive to the addition of 500 nM nanobody (black arrow, 250 sec) on supported lipid bilayers containing H-Ras(GTP) and farnesyl-Gβγ.

E. Kinetics of 50 nM DY647-p84/p110γ membrane recruitment appears indistinguishable in the absence and presence of nanobody. Prior to sample injection, DY647-p84/p110γ was incubated for 10 minutes with 500 nM nanobody.

F. Representative TIRF-M images showing the localization of 50 nM DY647-p84/p110γ visualized in the absence or presence of 500 nM nanobody (+NB7). Membrane composition for panels C-E: 93% DOPC, 5% DOPS, 2% MCC-PE, Ras(GTP) covalently attached to MCC-PE, and 200 nM farnesyl-Gβγ.

To further understand the mechanism by which this nanobody blocked lipid kinase activity we measured the bulk membrane recruitment dynamics of fluorescently labeled Dy647-p84-p110γ on supported lipid bilayers (SLBs) using Total Internal Reflection Fluorescence (TIRF) Microscopy. We found that the nanobody had no effect on membrane recruitment of p110γ-p84 on bilayers containing membrane-tethered Ras(GTP) and Gβγ (Fig. 1D-F). Membrane binding was not affected when the nanobody was spiked into samples containing membrane associated Dy647-p84-p110γ (Fig 1D). Similarly, pre-incubation of Dy647-p84-p110γ with 500 nM NB7 did not perturb membrane association of the kinase when flowed over a supported membrane (Fig. 1E).

We wanted to define the molecular mechanism of how nanobody NB7 was a potent allosteric inhibitor of lipid kinase activity. We purified the complex of nanobody NB7 bound to p110γ-p84 to homogeneity by gel filtration. Using this sample, we obtained a cryo-EM reconstruction of the complex of nanobody (NB7)-bound p110γ at 3.0-Å overall resolution from 149,603 particles (Figs. 2A-D, S1 and supplemental table 1). The density map was of sufficient quality to perform automated and manual construction of the p110γ-NB7 complex, with unambiguous building of the interfacial contacts between NB7 and p110γ. Nanobody binding did not induce any large-scale conformational changes of p110γ, as the structure of p110γ bound to NB7 was similar to the apo p110γ crystal structure or p110γ-p101 cryo-EM structure (Fig. S2). The lowest local resolution was in the ABD domain, with increased B-factors of the ABD in the p110γ-NB7 structure compared to p110γ-p101 (Fig. S2). This is consistent with the concept that ABD flexibility plays an important role in class I PI3K regulation (Liu et al., 2022).

Structure of p110γ bound to inhibitory nanobody NB7

A. Domain schematics of p110γ with helical domain (blue), activation loop (orange), and regulatory motif (green) of p110 annotated.

B. Density map of the p110γ-NB7 complex colored according to the schematic in A.

C. Cartoon model of the structure of p110γ bound to NB7 colored according to A.

D. Schematic depicting the key features of p110 and the nanobody binding site, colored according to panel A.

E. Domain schematic of NB7 CDR regions and their sequences.

F. Zoom in on the binding interface of NB7, with the CDRs colored as in panel E, and the electron density of the CDR regions contoured at 3σ (blue mesh).

The interface between NB7 and p110γ was extensive, with ∼1200 Å of buried surface area, with interactions of the ABD-RBD linker, N-terminus of the helical domain, and the regulatory motif at the turn between kα8-kα9 (1022-1026aa). This location in the regulatory motif is where both activating oncogenic (R1021C) and inhibitory loss of function mutants have been identified (R1021P) (Takeda et al., 2019), as well as a putative inhibitory phosphorylation site (T1024) (Perino et al., 2011). The resolution was sufficient to unambiguous build the three complementarity determining region (CDR) loops of NB7 that mediate target selectivity (Fig. 2E-F). The interface is primarily hydrophobic, with only 8 hydrogen bonds, and 1 electrostatic interaction among the 33 interfacial residues of NB7. A pocket formed between the helical domain and the ABD-RBD linker forms the majority of the interface, with extensive interactions with the long CDR3. The CDR1 loop packed up against the N-terminal section of the helical domain, with the CDR2 loop forming the interface with the regulatory motif. Previous study of oncogenic mutants in the regulatory motif of p110γ showed that increased dynamics mediated by these mutants increased kinase activity, putatively by breaking the autoinhibitory tryptophan lock in kα12 of the regulatory motif. Therefore, rigidifying the regulatory motif likely explains the molecular basis for how it prevents kinase activity. The nanobody interface is distinct from the predicted Gβγ interface (Rathinaswamy et al., 2023) and the experimentally resolved Ras interface (Pacold et al., 2000), explaining why it can still be membrane recruited by these stimuli.

p110γ activation by helical domain phosphorylation

To further understand the potential role helical domain dynamics in regulating p110γ activity we searched for any possible regulatory roles of this in PI3K signalling. It has been previously identified that p110γ is directly phosphorylated by the protein kinase PKCβ on residue S582 in the helical domain (Walser et al., 2013). S582 is located on the interior of the helical domain, and would not be expected to be exposed when the N-terminal region of the helical domain is folded (Fig. 3A). To understand this better at a molecular level, we purified a catalytic fragment of PKCβ and performed protein phosphorylation reactions on p110γ apo, p110γ-p84, and p110γ-p101. We identified a phosphorylated peptide containing S582, and surprisingly, we found an additional p110γ phosphorylation site at S594/S595 (Fig. 3B, Fig. S3). The S594/S595 site is also located in the N-terminal region of the helical domain, and is even more buried than S582, and would not be expected to be exposed when this region is folded (Fig. 3A). Dose response curves of PKCβ treatment were carried out for p110γ (Fig. 3C), p110γ-p84 (Fig. 3D), and p110γ-p101 (Fig. 3E). Both p110γ and p110γ-p84 showed similar dose response curves for PKCβ treatment, with similar curves for S582 and S594/S595. The p110γ-p101 complex was only very weakly phosphorylated, with <100-fold lower levels compared to p110γ and p110γ-p84 (Fig. 3E). This is consistent with the helical domain in p110γ being more rigid when bound to p101, compared to either bound to p84 or p110γ alone.

PKC leads to dual phosphorylation of internal sites in the helical domain, with selectivity for apo p110γ and p110γ-p84 over p110γ-p101.

A. Putative phosphorylation sites mapped on the structure of p110γ (PDB: 7MEZ) and cartoon schematic. The regions are colored based on domain schematics featured in Fig 2A.

B. Raw MS spectra of the unphosphorylated and phosphorylated peptide for a region spanning 579-592 (RYESLKHPKAYPKL) and 593-607 (FSSVKWGQQEIVAKT).

C-E. Extracted traces and ratios of the intensity of extracted ion traces of different phosphorylation site peptides (Top to bottom: S594/S595 and S582) from © p110γ, (D) p110γ/p84 or (E) p110γ/p101 samples treated with increasing concentration of PKCb according to the legend.

To provide additional insight into the molecular mechanisms underlying p110γ phosphorylation we carried out hydrogen deuterium exchange mass spectrometry (HDX-MS) experiments on p110γ and phosphorylated p110γ (90.8% phosphorylated S594/595, 92% phosphorylated S582) (Fig. 4A). The full data underlying the experiment is available in the source data, and data processing information is in supplemental table 2. We have previously observed that the N-terminal region of the helical domain of apo p110γ (residues spanning 557-630aa) shows isotope profiles that are consistent with EX1 H/D exchange kinetics (Rathinaswamy et al., 2021b, 2021c; Vadas et al., 2013; Walser et al., 2013). This is indicative of cooperative unfolding of extended protein regions, with H/D exchange occurring faster than the refolding event. This region is where the PKC phosphorylation sites are located, and may explain how the buried residues S582 and S594/S595 can be exposed to PKC. This is compatible with the observation that p110γ-p101 is protected from phosphorylation, as it does not show EX1 kinetics in this region, whereas both p110γ and p110γ-p84 do (Rathinaswamy et al., 2021a).

Activating phosphorylation at the helical domain leads to opening of the regulatory motif

A. HDX-MS comparing apo and phosphorylated p110γ. Significant differences in deuterium exchange are mapped on to the structure and cartoon of p110γ according to the legend (PDB: 7MEZ).

B. The graph of the #D difference in deuterium incorporation for p110γ, with each point representing a single peptide. Peptides colored in red are those that had a significant change in the mutants (greater than 0.4 Da and 5% difference at any timepoint, with a two tailed t-test p&lt;0.01). Error bars are S.D. (n=3).

C. Representative bimodal distribution (EX1 kinetics) observed in the helical domain peptides of p110γ.

D. Representative p100γ peptides displaying increases in exchange in the phosphorylated state are shown. For all panels, error bars show SD (n = 3)

E. Lipid kinase activity assays of phosphorylated and non-phosphorylated p100γ (concentration, 12nM to 1000nM) ATPase activity (left) and membrane (5% phosphatidylinositol 4,5-bisphosphate (PIP2), 95% phosphatidylserine (PS)) activation (right). Significance is indicated by **(&lt;0.001%), and ***(&lt;0.0001%).

When we compared phosphorylated p110γ (>90.8% as measured by mass spectrometry at both sites) to unphosphorylated p110γ we observed extensive increases in dynamics in the C2, helical domain and kinase domain (Fig. 4A-D). The largest increases in exchange upon phosphorylation were located in the N-terminal region of the helical domain, with the peptides directly adjacent to the phosphorylation site showing almost complete deuterium incorporation at the earliest time points of exchange. This is indicative of significant disruption of the alpha helical secondary structure in this region. When we examined the exchange profiles in this region they still underwent EX1 kinetics (Fig. 4C), however, phosphorylated p110γ was enriched in the more fully deuterated species. In addition to the regions in the helical domain, a portion of the regulatory motif of the kinase domain also showed increased deuterium exposure. This included the kα9-kα12 helices that surround the activation loop of p110γ. These increases in exchange were similar to those we had observed in a R1021C oncogenic activating mutant of PIK3CG (Rathinaswamy et al., 2021c).

To further explore the potential role of phosphorylation in mediating p110γ activity, we examined the kinase activity of p110γ for both basal ATPase and lipid membranes. Assays measured the hydrolysis of ATP to ADP, with this acting as a surrogate for PIP3 production in the membrane assay. PKCβ-mediated phosphorylation enhanced the ATPase activity of p110γ ∼2-fold in both the absence and presence of membranes (Fig. 4E).

Nanobody blocks p110γ phosphorylation

As NB7 bound at the interface of the helical and kinase domains that is exposed upon PKC phosphorylation of p110γ we hypothesized that the nanobody would likely alter phosphorylation. We carried out PKC phosphorylation of p110γ, p110γ-p101 and p110γ bound to NB7. The presence of NB7 showed even lower phosphorylation than p110γ-p101, with almost complete abrogation of phosphorylation at both sites (Fig. 5A-B). We also wanted to determine whether p110γ phosphorylation reciprocally perturbed NB7 binding. BLI experiments showed that there was no detectable binding of NB7 to phosphorylated p110γ (Fig. 5C-D), consistent with phosphorylation disrupting the N-terminal region of the p110γ helical domain. In addition, lipid kinase assays using phosphorylated p110γ showed no detectable difference in activity when measured in the absence and presence nanobody (Fig. 5E).

Nanobody NB7 blocks PKC phosphorylation, and phosphorylation prevents nanobody binding

A. Graph showing the intensities of phosphorylated and non-phosphorylated p110γ peptide (593-607) for PKC (500 nM) treated p110γ (black), PKC treated p110γ with NB7 (red) and PKC treated p110γp101 (purple). Scatter plot showing the percent phosphorylation of each complex from the left graph for the indicated peptide (n=3, right). Significance is indicated by ***(&lt;0.0001%).

B. Graph showing the intensities of phosphorylated and non-phosphorylated p110γ peptide (579-592) for PKC treated p110γ (black), PKC treated p110γ with NB7 (red) and PKC treated p110γp101 (purple). Scatter plot showing the percent phosphorylation of each complex from the left graph for the indicated peptide (n=3, right). Significance is indicated by * (&lt;0.01%), and ***(&lt;0.0001%).

C. Cartoon schematic of BLI analysis of the binding of immobilized His-NB7 to phosphorylated and non-phosphorylated p110γ.

D. Association curves for phosphorylated and non-phosphorylated p110γ (25nM) binding to His-NB7 are shown (n=3).

E. Kinase activity assays comparing the activation/inhibition of phosphorylated and non-phosphorylated p110γ (concentration, 12nM to 1000nM) with or without nanobody (3000nM final). Significance is indicated by * (&lt;0.05%), and NS (>0.05%).

Discussion

Here we find that the helical domain is a central regulator of the p110γ catalytic subunit of class IB PI3K, with modulation of helical dynamics through binding partners or PTMs able to either increase or decrease lipid kinase activity. These results expand on previous work defining the helical domain as a central regulator of class IA PI3Ks, where the nSH2 domain of the p85 regulatory subunits makes inhibitory interactions that significantly inhibit lipid kinase activity of all class IA catalytic subunits (p110α, p110β, and p110δ) (Mandelker et al., 2009; Miled et al., 2007; Burke and Williams, 2013; Burke, 2018). This inhibitory interaction in class IA PI3Ks is disrupted in human cancers (helical hotspot mutations in PIK3CA) (Samuels et al., 2004) and immune disorders (helical mutations in PIK3CD in APDS1) (Angulo et al., 2013; Lucas et al., 2014). Class IB PI3Ks are unique compared to class IA PI3Ks, as they are not inhibited by p101 and p84 regulatory subunits, but instead potentiate GPCR activation. This lack of inhibition is due to the distinct binding interface of class IB PI3K regulatory subunits compared to class IA regulatory subunits, with only class IA regulatory subunits making direct inhibitory interactions with the kinase and helical domains of p110 catalytic subunits (Rathinaswamy et al., 2021a). Here we show that a unique surface at the interface of the helical and kinase domains of p110γ is a potential site for the development of novel allosteric inhibitors that modulate p110γ activity.

The previously identified inhibitory nanobody (NB7) (Rathinaswamy et al., 2021b) bound with high affinity and inhibited all complexes of p110γ. The nanobody interface is distinct from how the nSH2 inhibits class IA PI3K activity, as its binding site is on the opposite face of the helical domain (Fig. 6A). The mechanism of inhibition is also distinct, as the nSH2-helical interaction plays a critical role in preventing membrane recruitment of inhibited class I PI3Ks, with removal of this interface either through pYXXM motif binding, or oncogenic mutations leading to increased membrane recruitment (Burke et al., 2012, 2011; Zhang et al., 2011). Analysis of the nanobody binding site compared to the structure of HRas-p110γ or the HDX-MS supported Alphafold-multimer prediction of Gβγ-p110γ (Rathinaswamy et al., 2023) shows that nanobody binding does not sterically block complex formation (Fig. 6B). This is consistent with it not blocking membrane recruitment by Ras/Gβγ.The nanobody inhibited ATP turnover both in solution and on membranes, suggesting that it prevents formation of a catalytically competent conformation of p110γ, but still allows for membrane recruitment. Further development of small molecule allosteric binding partners in this allosteric pocket between the kinase and helical domain may reveal the specific molecular interactions in this pocket that mediate inhibition.

Comparison of nanobody binding site compared to p85 inhibition of class IA PI3Ks and class IB activation sites

A. Comparison of the nanobody NB7binding site in p110γ compared to the nSH2 inhibitory site in p110((PDB: 3HHM) (Mandelker et al., 2009)

B. Comparison of the nanobody NB7binding site in p110γ compared to the X-ray structure of the Ras binding site (PDB: 1HE8) (Pacold et al., 2000) and the Alphafold model of Gβγ bound to p110γ (Rathinaswamy et al., 2023).

C. Oncogenic mutations and post-translational modifications in spatial proximity to the nanobody binding site.

Oncogenic mutations are frequent in the class IA PI3Kα encoded by PIK3CA, with this being the 2nd most frequently mutated gene in human cancer (Lawrence et al., 2014). Mutations in p110γ encoded by PIK3CG in cancer are less frequent, however, they can still provide insight into regulatory mechanisms that control activity. Oncogenic mutations in the kinase domain (R1021C) and helical domain (E581K) are in close proximity to the nanobody binding site, and both would be expected to disrupt the stability of the helical domain or regulatory motif of the kinase domain (Fig. 6C). In addition to these mutations there are also multiple post-translational modifications that occur in this region, including inhibitory phosphorylation at T1024 (Perino et al., 2011), and activating phosphorylation at S582 (Walser et al., 2013). PKCβ is activated downstream of the IgE receptor in mast cells (Walser et al., 2013), but the full details of how this activates PI3K has been unclear. We identified an additional PKCβ phosphorylation site located in the helical domain (S594/S595) (Fig. 6C). Both the S582 and S594/S595 sites are not surface accessible and would require a transient opening of the helical domain for kinase accessibility. HDX-MS analysis of the helical domain of p110γ has shown that it is more dynamic than other class I PI3K isoforms (Burke and Williams, 2013; Walser et al., 2013), with the presence of the p101 regulatory subunit dramatically decreasing helical domain dynamics (Vadas et al., 2013). This putative mechanism of helical domain dynamics driving PKCβ phosphorylation is consistent with our observation that p101 subunits decreased p110γ phosphorylation >100-fold. PKCβ phosphorylation of p110γ leads to increased dynamics in both the helical and kinase domains with increased kinase activity, although only weakly compared to full activation by either membrane localised Ras or Gβγ. This increase was observed with both membrane and soluble substrate, so likely is not driven by altered membrane recruitment.

Overall, our biophysical and biochemical analysis of modulators of helical domain dynamics reveal the critical role of this domain in regulating class IB PI3K activity. This raises possibilities for development of small molecule modulators that may either increase or decrease helical domain dynamics, leading to either activation or inhibition. The high-resolution structure of an allosteric inhibitor nanobody provide initial insight into which pockets can specifically be targeted. Multiple ATP competitive p110γ selective inhibitors are in clinical trials for human cancers (Li et al., 2021), with many having significant side effects. The identification of novel inhibitory strategies provides new opportunities for targeting p110γ dysregulation in human disease.

eLife Assessment

This study presents fundamental new insight into the regulatory apparatus of PI3Kgamma, a crucial kinase in signaling pathways that control the immune response and cancer. A suite of biophysical and biochemical approaches provide convincing evidence for new sites of allosteric control over enzyme activity. The rigorous findings provide structure and dynamic information that may be exploited in efforts to control PI3Kgamma activity in a therapeutic setting.

Reviewer #1 (Public Review):

The authors examine signaling factors that differentiate parallel routes to activating phosphoinositide 3-kinase gamma (PI3Kγ). Dissecting the convergent pathways that control PI3Kγ activity is critical because PI3Kγ is a therapeutic target for treating inflammatory disease and cancer. Here, the authors employ a multipronged approach to reveal new aspects for how p84 and p101 pair with p110γ to activate the PI3Kγ heterodimer. The key instigator to this study is a previously reported inhibitory Nanobody, NB7. The hypothesized mechanism for NB7 allosteric inhibition of p84- p110γ was previously proposed to involve blockage of the Ras-binding domain. The authors revise the allosteric inhibition model based on meticulous profiling of various PI3Kγ complex interactions with NB7. In parallel, a cryo-EM-derived model of NB7 bound to the p110γ subunit convincingly reveals a Nanobody interaction pocket involving the helical domain and regulatory motifs of the kinase domain. This revelation shifts the focus to the helical domain, a known target of PKC phosphorylation. While the connections between NB7 interactions and the effects of PKC phosphorylation are sometimes tenuous, it could be argued that the Nanobody served as a tool to reveal the importance of the helical domain to p110γ regulation.

The sites of PKC-mediated p110γ helical domain phosphorylation were unexpectedly inaccessible in the available structural models. Nevertheless, mass spectrometry (MS)-based phosphorylation profiling indicates that PKC can phosphorylate the helical domain of p110γ and p84/p110γ (but not p101/p110γ) in vitro. The authors hypothesize that helical domain dynamics dictate susceptibility to PKC phosphorylation. To explore this notion, carefully executed, rigorous H/D exchange MS (HDX-MS) experiments were performed comparing phosphorylated vs. unphosphorylated p110γ. Notably, this design reveals more about the consequences of p110γ phosphorylation, rather than the mechanisms of p84/p101 promoting/resisting phosphorylation. Nevertheless, HDX-MS is very well suited to exploring secondary structure dynamics, and helical domain phosphorylation strikingly increases dynamics consistent with increased regional accessibility. The increased dynamics also nicely map to the pocket enveloped by the inhibitory NB7 Nanobody.

Ultimately, this study reveals an unexpected p110γ pocket that allows an engineered Nanobody to allosterically inhibit PI3Kγ complexes. The cryo-EM characterization of the interaction inspired an HDX-MS investigation of known sites of phosphorylation in the region. These insights could be linked to differences/convergences of p84 and p101 complex formation and activation of PI3Kγ, and future work may clarify these mechanisms further. The data presented herein will also be useful for broadening the target surface for future therapeutic developments. New allosteric connections between effector binding sites and post-translational modifications are always welcome.

Reviewer #2 (Public Review):

Harris et al. have described the cryo-EM structure of PI3K p110gamma in a complex with a nanobody that inhibits the enzyme. This provided the first structure of full-length of PI3Kgamma in the absence of a regulatory subunit. This nanobody is a potent allosteric inhibitor of the enzyme, and might provide a starting point for developing allosteric, isotype-specific inhibitors of the enzyme. One distinct effect of the nanobody is to greatly decrease the dynamics of the enzyme as shown by HDX-MS, which is consistent with a growing body of observations suggesting that for the whole PI3K superfamily, enzyme activators increase enzyme dynamics.

The most remarkable outcome of the study is that upon observing the site of nanobody binding, the authors searched the literature and found that there was a previous report of a PKCbeta phosphorylation of PI3Kgamma in the helical domain that is near the nanobody binding site. This led the authors to re-examine the consequence of the phosphorylation armed with better structural models and the tools to study the effects of this phosphorylation on enzyme dynamics. They found that the site of phosphorylation is buried in the helical domain, suggesting that a large conformational change would have to take place to enable the phosphorylation. HDX-MS showed that phosphorylation at three sites clustered in the helical domain generate a distinctly different conformation with rapid deuterium exchange. This suggests that the phosphorylation locks the enzyme in a more dynamic state. Their enzyme kinetics show that the phosphorylated, dynamic enzyme is activated.

While this phosphorylation was reported before, the authors have provided a mechanism for why this activates the enzyme, and they have shown why binders that stabilise the helical domain (such as binding to the p101 regulatory subunit and the nanobody) prevent the phosphorylation. It is this insight into the dynamics of the PI3Kgamma that will likely be the long-lasting influence of the work.

The paper is well written and the methods are clear.

Author Response:

The following is the authors' response to the original reviews.

We appreciate the in depth review of our manuscript, and the excellent suggestions from the two reviewers. We have addressed all concerns as described in the point by point response below. We have also added all of these changes to a revised version posted to biorxiv on May 23rd 2023 (BIORXIV/2023/536585).

Reviewer #1 (Recommendations For The Authors):

It is sometimes difficult to connect the rationalizations behind the transitions between NB7 binding interaction, the compare/contrast of p84 and p101 effectors, and the synergy with phosphorylation. More explanation of the rationalizations behind these transitions in the Results would be helpful.

We agree that the manuscript would benefit from better transitions between the sections. We have added a new paragraph in the final section describing the nanobody structure before the helical domain phosphorylation that fully describes the rationale for how both inform on the critical role of helical domain dynamics in kinase activity. This paragraph is shown below.

_‘The interface of NB7 with p110_g _is distant from both the putative membrane binding surface, as well as the catalytic machinery of the kinase domain. To further understand how this nanobody could so potently inhibit PI3K activity we examined any other potential modulators of PI3K activity localised in this region. There are two regulatory phosphorylation sites in the helical (Walser et al., 2013) and kinase domain (Perino et al., 2011) localised at the NB7 interface. This is intriguing as helical domain phosphorylation is activating, and kinase domain phosphorylation is inhibitory. This suggested a critical role in the regulation of p110_g is the dynamics of this kinase-helical interface. To fully define the role of NB7 in altering the dynamics of the helical domain we needed to study other modulators of helical domain dynamics.’

The Methods section would benefit from careful copy editing for clarity and consistency.

We have gone through the methods section and edited for clarity and consistency throughout.

There's a minor ambiguity throughout when referring to the phosphorylation of S594/S595. Although close inspection makes it clear that this refers to the monophosphorylation of either site S594 or S595, there are several references to "S594/S595" that could be interpreted as phosphorylation of both residues.

We agree that this was ambiguous in the original text. We have added an explicit statement describing this as a single phosphorylation event.

‘The modification at this site results in a single phosphorylation event , but due to CID MS/MS fragmentation we cannot determine which site is modified, and will be described as S594/S595 throughout the manuscript.’

In Figure 2B, the authors show the cryo-EM density map and the structural model based on this map. It would be helpful to also include an image of the structural model fit in the density map to allow readers to evaluate the quality of the map and model. The 2F panel provides an important view of this fit, but CD3 models are difficult to discern.

We agree that this would help interpret model quality. We have added a new supplemental figure showing the fit of both p110 and NB7 into this Cryo EM density (see new Fig S2).

Paragraph starting at line 258: The shift to monitoring ATPase activity is confusing here. ATPase activity indicates production of ADP + phosphate (rather than ADP + PIP3). However, an explanation is provided that states that measuring ADP production serves as a surrogate for measuring PIP3 production. The apparent absence of membrane PIP2 substrate in Figure 4E (left) suggests that there is a true ATPase background activity in this kinase. If so, does the increase in ADP production in Fig. 4E reflect the inclusion of PIP2 substrate, increased background ATPase activity, or both?

We agree that this was worded confusingly in the original version. We have now clarified exactly what we are observing in these ATPase assays. The new paragraph is appended below

‘To further explore the potential role of phosphorylation in mediating p110g activity, we examined the kinase activity of p110g under two conditions: basal ATP turnover, and with PIP2 containing lipid membranes. The experiments in the absence of PIP2 measure turnover of ATP into ADP and phosphate, and is a readout of basal catalytic competency. Experiments with PIP2, measured ATP consumed in the generation of PIP3, as well as in non-productive ATP turnover. The p110g enzyme in the absence of stimulators is very weakly active towards PIP2 substrate with only ~2 fold increased ATP turnover compared to in the absence of membranes. This is consistent with very weak membrane recruitment of p110g complexes in the absence of lipid activators (Rathinaswamy et al., 2023). PKCb-mediated phosphorylation enhanced the ATPase activity of p110g ~2-fold in both the absence and presence of membranes (Fig. 4E). This suggests that the effect of phosphorylation is to change the intrinsic catalytic efficiency of phosphorylated p110g, with limited effect on membrane binding.’

In the section "Nanobody blocks p110gamma phosphorylation," it's not entirely convincing that "the presence of NB7 showed even lower phosphorylation than p110gamma-p101." This does not seem to be subject to a significance test in Figure 5A/B. The follow-up point about "complete abrogation of phosphorylation," however, is readily apparent.

We agree that we could have been more precise with our language, as this is not a complete block of phosphorylation, it is merely a significant decrease in phosphorylation. We have removed the comparison to p110-p101, and also removed the statement about complete abrogation of phosphorylation. This is now reworded to

‘The presence of NB7 showed a significant decrease in p110g phosphorylation at both sites (Fig. 5A-B).’

Figure 1: Legend needs to include more detail to define the data. (1A) Representations of variance need to be clarified (e.g., replicates, error bar meaning). Consider "Normalized lipid kinase activity" as a y-axis label and expand on the activity measurement and normalization in the legend. (1B) How was error calculated? (1C) Mislabeled as 1B? Also, consider clarifying the first title highlighting the comparison to class IA PI3Ks. (1D) Typo: "Y647-p84/p110gamma." Also, would it not be more accurate to say "effect of nanobody NB7 on PI3K displacement..." for this experiment?

We apologise for these oversights. See details on what has been changed in Fig 1A, 1B, 1C, and 1D.

For Fig 1A we now show the data where each replicate is indicated in the graph in the absence of error bars, and have also more clearly expanded on this activity measurement in the figure legend and also stated the replicate number.

For Fig 1B we now clearly state how the error was generated.

For Fig 1C we have fixed the typo

For Fig 1D, we have fixed this typo and also changed the sub-heading as suggested.

New figure legend is below as well

Figure 1. The inhibitory nanobody NB7 binds tightly to all p110γ complexes and inhibits kinase activity, but does not prevent membrane binding

A. Cartoon schematic depicting nanobody inhibition of activation by lipidated Gβγ (1.5 µM final concentration). Lipid kinase assays show a potent inhibition of lipid kinase activity with increasing concentrations of NB7 (3-3000 nM) for the different complexes. Experiments are carried out in triplicate (n=3) with each replicate shown. The y-axis shows lipid kinase activity normalised for each complex activated by Gβγ in the absence of nanobody. Concentrations of each protein were selected to give a lipid kinase value in the detectable range of the ATPase transcreener assay. The protein concentration of p110γ (300 nM), p110γ-p84 (330 nM) and p110γ-p101 (12 nM) was different due to intrinsic differences of each complex to be activated by lipidated Gβγ, and is likely mainly dependent for the difference seen in NB7 response.

B. Association and dissociation curves for the dose response of His-NB7 binding to p110γ, p110γ-p84 and p110γ-p101 (50 – 1.9 nM) is shown. A cartoon schematic of BLI analysis of the binding of immobilized His-NB7 to p110γ is shown on the left. Dissociation constants (KD) were calculated based on a global fit to a 1:1 model for the top three concentrations and averaged with error shown. Error was calculated from the association and dissociation value (n=3) with standard deviation shown. Full details are present in the source data.

C. Association and dissociation curves for His-NB7 binding to p110γ, p110a-p85a, p110b-p85b, and p110d-p85b. Experiments were performed in duplicate with a final concentration of 50 nM of each class I PI3K complex.

D. Effect of NB7 on PI3K recruitment to supported lipid bilayers containing H-Ras(GTP) and farnesyl-Gbg as measured by Total Internal Reflection Fluorescence Microscopy (TIRF-M). DY647-p84/p110g displays rapid equilibration kinetics and is insensitive to the addition of 500 nM nanobody (black arrow, 250 sec) on supported lipid bilayers containing H-Ras(GTP) and farnesyl-Gbg.

E. Kinetics of 50 nM DY647-p84/p110g membrane recruitment appears indistinguishable in the absence and presence of nanobody. Prior to sample injection, DY647-p84/p110g was incubated for 10 minutes with 500 nM nanobody.

F. Representative TIRF-M images showing the localization of 50 nM DY647-p84/p110g visualized in the absence or presence of 500 nM nanobody (+NB7). Membrane composition for panels C-E: 93% DOPC, 5% DOPS, 2% MCC-PE, Ras(GTP) covalently attached to MCC-PE, and 200 nM farnesyl-Gbg.

Figure 2: (1A) For consistency with the rest of the paper, p110g can be updated with the Greek character. (1B) This may have been intentional to attract attention to subdomains interacting with NB7, but "colored according to the schematic" omits the purple RBD. (2F) the figure legend should specify whether p110gamma surfaces depicted are the cryo-EM density or a surface rendition of the structural model.

We agree and have fixed the p110 typo, and have also colored the schematic the same as shown in the cartoon model.

The data shown in Fig 1B is indeed the Cryo EM density and this is now clearly indicated in the legend.

Figure 3: (3B) Specifying the [M+H] as [M+2H]2+ and [M+4H]4+ would help the reader understand the delta mass for monophosphorylation here. Given the broad readership of this journal, it would be useful to define 't' and 'e' as 'theoretical' and 'experimental' in the legend. It may also help to be explicit about the meaning of the red spectra and residues in the legend. (3C-E) autocorrect typo for "(C)" and an opportunity to update "b" for Greek character beta.

We agree that clearly defining the charge state of each spectra will make it more obvious that we are dealing with a mono-phosphorylation and have made this change as suggested in the figure. We have also clearly define m/z t and m/z e in the figure legend, as well as the black and red lines, and characters. Finally we have added PKCb for all descriptions of PKC treatment in the figure, and fixed the incorrect PKC’b’ in the legend.

Figure 4: (4C) Given the common use of "ND" for other terms, it would be useful to spell out "no deuterium" or "undeuterated." (4E) the parenthetical "(concentration, 12nM to 1000nM)" could be clarified. How are the (presumably p100gamma) concentration ranges reflected in the three plotted data points per treatment? See also Figure 5E.

We agree and have redefined ND as undeuterated. We apologise for the typo in the figure legend, as the concentrations of p110 gamma were the same for both phosphorylated and non-phosphorylated, with this being a typo (all concentrations of enzyme were 1000 nM). This has been changed here and in Fig 5E.

Figure 5: (5A/B) Some clarification that we're looking at extracted ion chromatograms would be very useful in this figure legend. On a related note, the experimental details on the LC-MS methodology for this data appear to be split between two sections of the Methods: the "Phosphorylation analysis" paragraph (line 526) and the HDX-MS section. Some more explicit cross-referencing would clarify this experiment. (5E) Clarify inclusion of PIP2 membranes here.

We have clearly described that we are looking at extracted ion chromatograms in both panel A and B. We also have normalized the experimental methods in the LC-MS as these used exactly the same procedure. Finally, we now clearly describe the assays shown in Fig 5E were performed in the absence of PIP2 membranes.

Miscellaneous typos:

Line 205: reference omitted for "Previous study.."

We have added this reference

Line 196: "unambiguous"

Fixed to unambiguously

Reviewer #2 (Recommendations For The Authors):

The only mistake I spotted was that on line 729 there is a reference to Fig 3C that should actually be Fig. 4C

We have changed this to the correct Fig 4C.

Version 2

Abstract

Abstract

PI3Kγ is a critical immune signaling enzyme activated downstream of diverse cell surface molecules, including Ras, PKCβ activated by the IgE receptor, and Gβγ subunits released from activated GPCRs. PI3Kγ can form two distinct complexes, with the p110γ catalytic subunit binding to either a p101 or p84 regulatory subunit, with these complexes being differentially activated by upstream stimuli. Here using a combination of cryo electron microscopy, HDX-MS, and biochemical assays we have identified novel roles of the helical domain of p110γ in regulating lipid kinase activity of distinct PI3Kγ complexes. We defined the molecular basis for how an allosteric inhibitory nanobody potently inhibits kinase activity through rigidifying the helical domain and regulatory motif of the kinase domain. The nanobody did not block either p110γ membrane recruitment or Ras/Gβγ binding, but instead decreased ATP turnover. We also identified that p110γ can be activated by dual PKCβ helical domain phosphorylation leading to partial unfolding of an N-terminal region of the helical domain. PKCβ phosphorylation is selective for p110γ-p84 compared to p110γ-p101, driven by differential dynamics of the helical domain of these different complexes. Nanobody binding prevented PKCβ mediated phosphorylation. Overall, this works shows an unexpected allosteric regulatory role of the helical domain of p110γ that is distinct between p110γ-p84 and p110γ-p101 and reveals how this can be modulated by either phosphorylation or allosteric inhibitory binding partners. This opens possibilities of future allosteric inhibitor development for therapeutic intervention.

Introduction

The class I phosphoinositide 3 kinases (PI3Ks) are master regulators of myriad functions through their generation of the lipid signalling molecule phosphatidylinositol 3,4,5-trisphosphate (PIP3) downstream of cell surface receptors (Burke and Williams, 2015; Rathinaswamy and Burke, 2019; Vanhaesebroeck et al., 2021; Vasan and Cantley, 2022). The class I PI3Ks can be further subdivided into the class IA and class IB subfamilies, with class IB PI3Ks being critical in immune signalling, and are composed of a single p110γ catalytic subunit that can bind to either a p101 or p84 regulatory subunit (Hawkins and Stephens, 2015; Lanahan et al., 2022; Okkenhaug, 2013). The two PI3Kγ complexes (either p110γ-p84 or p110γ-p101) play essential and independent roles in both the adaptative and innate immune systems. PI3Kγ has shown promise as a therapeutic target, primarily as an immunomodulator of the tumor microenvironment leading to enhanced anti-tumor immune responses (De Henau et al., 2016; Kaneda et al., 2016b). Multiple isoform selective small molecule ATP competitive inhibitors of p110γ are in clinical trials for multiple forms of human cancers (Li et al., 2021). However, all inhibitors currently developed towards p110γ act as ATP competitive inhibitors, showing equal potency against both p110γ-p84 or p110γ-p101 complexes.

Detailed experiments on the role of p110γ in mice show that knockout of both p101 and p84 leads to PIP3 levels that are equivalent to knockout of p110γ, showing that all cellular PI3Kγ activity requires the presence of either a p84 or p101 regulatory subunit (Rynkiewicz et al., 2020). The two complexes are differentially activated by membrane localised receptors, including G-protein coupled receptors (Li et al., 2000; Stephens et al., 1997), Ras (Jin et al., 2020; Kurig et al., 2009), toll like receptors (TLRs) (Luo et al., 2018), and the IgE antigen receptor (Laffargue et al., 2002; Walser et al., 2013). This leads to the different complexes driving unique immune responses, with p110γ-p101involved in chemotaxis in neutrophils (Bohnacker et al., 2009; Deladeriere et al., 2015), and p110γ-p84 involved in reactive oxide production. Differential activation of unique PI3Kγ complexes downstream of GPCRs and Ras is caused by the ability of p101 to directly bind to Gβγ subunits downstream of activated GPCRs, with this being lost in p84, making p110γ-p84 activation by Gβγ dependent on Ras mediated membrane recruitment (Rathinaswamy et al., 2023; Kurig et al., 2009; Rynkiewicz et al., 2020). Activation of PI3Kγ downstream of the IgE antigen receptor is driven by calcium mediated activation of protein kinase Cβ, leading to the selective phosphorylation and activation of p110γ at S582 (Walser et al., 2013), with this putatively only occurring in p110γ-p84 and not p110γ-p101. The full molecular mechanisms underlying how phosphorylation of p110γ is selective for different p84 or p101 complexes, and how it activates lipid kinase activity are poorly understood.

Extensive biophysical approaches including cryo electron microscopy (cryo-EM), X-ray crystallography, and hydrogen deuterium exchange mass spectrometry (HDX-MS) have provided extensive insight into the molecular underpinnings of how p110γ associates with both p101 and p84, how they are differentially activated by Ras and GPCR signals, and how they can be activated on lipid membranes (Pacold et al., 2000; Walker et al., 1999; Rathinaswamy et al., 2021c, 2021a; Gangadhara et al., 2019; Vadas et al., 2013; Rathinaswamy et al., 2023). The p110γ catalytic subunit is composed of an adaptor binding domain (ABD), a Ras binding domain (RBD), a C2 domain, a helical domain, and a bi-lobal kinase domain (Rathinaswamy et al., 2021a; Walker et al., 1999). A set of helices positioned C-terminal to the activation loop in the kinase domain play a critical role in regulating activity, with this region referred to as the regulatory motif (Rathinaswamy et al., 2021c). The p110γ isoform is unique in that it is inhibited in the absence of a regulatory subunit, with this driven by an autoinhibitory conformation of the regulatory motif, that is proposed to require membrane association to disrupt (Gangadhara et al., 2019). The regulatory motif is a common site of activating mutations in the other class I PI3K isoforms (Jenkins et al., 2023), with p110γ having rare activating oncogenic mutants in this region (Rathinaswamy et al., 2021c). The p110γ subunit interacts with both p84 and p101 at an interface composed of the C2 domain, and the linkers between the RBD-C2 and C2-helical domains (Rathinaswamy et al., 2023, 2021a). The p110γ-p84 complex forms a more dynamic complex compared to p110γ-p101 (Rathinaswamy et al., 2023; Shymanets et al., 2013), however, no clear unique regulatory role of this difference in dynamics has been identified.

The fundamental roles of p110γ in inflammatory processes has made it a therapeutic target in many pathological conditions, including asthma (Campa et al., 2018), arthritis (Camps et al., 2005), obesity (Becattini et al., 2011; Breasson et al., 2017), and cancer (De Henau et al., 2016; Kaneda et al., 2016a, 2016b). There are significant side effects from non-isoform selective PI3K inhibitors (Bohnacker et al., 2017; Vanhaesebroeck et al., 2021), which has driven the development of highly p110γ selective small molecule inhibitors (Bell et al., 2012; Evans et al., 2016; Gangadhara et al., 2019). However, all p110γ inhibitors will target both p110γ-p101 and p110γ-p84, so there is a potential for the development of allosteric inhibitors outside of the ATP binding site. Initial promise has been reported for the development of class IA p110α oncogene specific allosteric inhibitors. However, further investigation of the molecular mechanisms underlying p110γ regulation will be required for the discovery of regions that can be targeted for allosteric inhibitor development.

Here we report critical roles of the helical domain of p110γ in both activation and inhibition of lipid kinase activity. We characterised an allosteric inhibitory nanobody (NB7) that potently inhibits p110γ activity. Cryo-EM was used to define the inhibitory interface, which is composed of the helical domain, the ABD-RBD linker, and the regulatory motif of the kinase domain of p110γ. The region that the nanobody binds to is in close spatial proximity to a previously identified PKCβ phosphorylation site (S582) in the helical domain, and oncogenic activating mutants in the regulatory motif. We fully characterised the activity and dynamics of stoichiometrically PKCβ phosphorylated p110γ, leading to the discovery of a novel additional phosphorylation site (either S594 or S595). PKCβ phosphorylation was highly selective for p110γ and p110γ-p84, with limited phosphorylation of p110γ-p101. Hydrogen deuterium exchange mass spectrometry (HDX-MS) analysis showed that phosphorylation of p110γ leads to unfolding of the N-terminal region of the helical domain, and increased kinase activity. The presence of the inhibitory nanobody significantly blocks PKCβ phosphorylation, while phosphorylation of p110γ prevented binding to NB7. Overall, this work provides unique insight into the critical role of the helical domain in controlling p110γ activity, and how phosphorylation and binding partners can modify this regulation. It also reveals a unique binding site located at the interface of the helical and kinase domain that can be targeted for future allosteric inhibitor design.

Results

Molecular mechanism of nanobody inhibition of p110γ

We previously identified multiple nanobodies that inhibited the activity of p110γ. One from this group (denoted NB7 throughout the manuscript) potently inhibited the membrane mediated activation of p110γ-p84 by both Ras and Gβγ, with HDX-MS experiments mapping the NB7 binding interface to the RBD, helical and kinase domains (Rathinaswamy et al., 2021b). We originally hypothesized that NB7 worked by sterically inhibiting Ras binding to the RBD domain of p110γ. To further explore the molecular mechanism of inhibition we purified all complexes of p110γ (p110γ apo, p110γ-p84, p110γ-p101) along with the NB7 nanobody. The SDS-PAGE of all proteins utilised in this study are shown in the source data file included in the supplemental information.

To define the mechanism concerning how NB7 inhibits PI3K activity we analyzed how this nanobody inhibited all class IB PI3K complexes (p110γ, p110γ-p84, p110γ-p101) upon activation by lipidated Gβγ subunits. Intriguingly, we found that all three forms of p110γ were potently inhibited by NB7 (Fig. 1A). While the IC50 measured for the three complexes was different, this is likely mainly due to the dramatic difference in protein required to measure lipid kinase activity in vitro (∼300 nM for p110γ apo/p110γ-p84, and ∼10 nM for p110γ-p101, respectively). This suggested that the mechanism of inhibition was not driven by a steric block of Ras association through the RBD, as previously proposed (Rathinaswamy et al., 2021b). We examined the binding of this nanobody to all complexes using biolayer interferometry (BLI). The nanobody bound equivalently tightly to all complexes, with ∼2 nM potency for p110γ, p110γ-p84, and p110γ-p101 (Fig. 1B). We also tested binding of the nanobody to all class IA PI3Ks, and there was no detectable binding to p110α, p110β, and p110δ (Fig. 1C).

The inhibitory nanobody NB7 binds tightly to all p110γ complexes and inhibits kinase activity, but does not prevent membrane binding

A. Cartoon schematic depicting nanobody inhibition of activation by lipidated Gβγ (1.5 µM final concentration) on 5% PIP2 membrane (5% phosphatidylinositol 4,5-bisphosphate (PIP2), 95% phosphatidylserine (PS)) activation. Lipid kinase assays showed a potent inhibition of lipid kinase activity with increasing concentrations of NB7 (3-3000 nM) for the different complexes. Experiments are carried out in triplicate (n=3) with each replicate shown. The y-axis shows lipid kinase activity normalised for each complex activated by Gβγ in the absence of nanobody. Concentrations of each protein were selected to give a lipid kinase value in the detectable range of the ATPase transcreener assay. The protein concentration of p110γ (300 nM), p110γ-p84 (330 nM) and p110γ-p101 (12 nM) was different due to intrinsic differences of each complex to be activated by lipidated Gβγ and is likely mainly dependent for the difference seen in NB7 response.

B. Association and dissociation curves for the dose response of His-NB7 binding to p110γ, p110γ-p84 and p110γ-p101 (50 – 1.9 nM) is shown. A cartoon schematic of BLI analysis of the binding of immobilized His-NB7 to p110γ is shown on the left. Dissociation constants (KD) were calculated based on a global fit to a 1:1 model for the top three concentrations and averaged with error shown. Error was calculated from the association and dissociation value (n=3) with standard deviation shown. Full details are present in the source data.

C. Association and dissociation curves for His-NB7 binding to p110γ, p110α-p85α, p110β-p85β, and p110δ-p85β. Experiments were performed in duplicate with a final concentration of 50 nM of each class I PI3K complex.

D. Effect of NB7 on PI3K recruitment to supported lipid bilayers containing H-Ras (GTP) and farnesyl-Gβγ as measured by Total Internal Reflection Fluorescence Microscopy (TIRF-M). DY647-p84/p110γ displays rapid equilibration kinetics and is insensitive to the addition of 500 nM nanobody (black arrow, 250 sec) on supported lipid bilayers containing H-Ras (GTP) and farnesyl-Gβγ.

E. Kinetics of 50 nM DY647-p84/p110γ membrane recruitment appears indistinguishable in the absence and presence of nanobody. Prior to sample injection, DY647-p84/p110γ was incubated for 10 minutes with 500 nM nanobody.

F. Representative TIRF-M images showing the localization of 50 nM DY647-p84/p110γ visualized in the absence or presence of 500 nM nanobody (+NB7). Membrane composition for panels C-E: 93% DOPC, 5% DOPS, 2% MCC-PE, Ras (GTP) covalently attached to MCC-PE, and 200 nM farnesyl-Gβγ.

To further understand the mechanism by which this nanobody blocked lipid kinase activity we measured the bulk membrane recruitment dynamics of fluorescently labeled Dy647-p84-p110γ on supported lipid bilayers (SLBs) using Total Internal Reflection Fluorescence (TIRF) Microscopy. We found that the nanobody had no effect on membrane recruitment of p110γ-p84 on bilayers containing membrane-tethered Ras (GTP) and Gβγ (Fig. 1D-F). Membrane binding was not affected when the nanobody was spiked into samples containing membrane associated Dy647-p84-p110γ (Fig 1D). Similarly, pre-incubation of Dy647-p84-p110γ with 500 nM NB7 did not perturb membrane association of the kinase when flowed over a supported membrane (Fig. 1E).

We wanted to define the molecular mechanism of how nanobody NB7 was a potent allosteric inhibitor of lipid kinase activity. We purified the complex of nanobody NB7 bound to p110γ-p84 to homogeneity by gel filtration. Using this sample, we obtained a cryo-EM reconstruction of the complex of nanobody (NB7)-bound p110γ at 3.0-Å overall resolution from 149,603 particles (Figs. 2A-D, S1-S2 and supplemental table 1). The density map was of sufficient quality to perform automated and manual construction of the p110γ-NB7 complex, with unambiguous building of the interfacial contacts between NB7 and p110γ (Fig. S2). Nanobody binding did not induce any large-scale conformational changes of p110γ, as the structure of p110γ bound to NB7 was similar to the apo p110γ crystal structure or p110γ-p101 cryo-EM structure (Fig. S3). The lowest local resolution was in the ABD domain, with increased B-factors of the ABD in the p110γ-NB7 structure compared to p110γ-p101 (Fig. S3). This is consistent with the concept that ABD flexibility plays an important role in class I PI3K regulation (Liu et al., 2022).

Structure of p110γ bound to inhibitory nanobody NB7

A. Domain schematics of p110γ with helical domain (blue), activation loop (orange), and regulatory motif (green) of p110γ annotated.

B. Cryo EM density of the p110γ-NB7 complex colored according to the schematic in A.

C. Cartoon model of the structure of p110γ bound to NB7 colored according to A.

D. Schematic depicting the key features of p110 and the nanobody binding site, colored according to panel A.

E. Domain schematic of NB7 CDR regions and their sequences.

F. Zoom in on the binding interface of NB7, with the CDRs colored as in panel E, and the electron density of the CDR regions contoured at 3σ (blue mesh).

The interface between NB7 and p110γ was extensive, with ∼1200 Å of buried surface area, with interactions of the ABD-RBD linker, N-terminus of the helical domain, and the regulatory motif at the turn between kα8-kα9 (1022-1026aa). This location in the regulatory motif is where both activating oncogenic (R1021C) and inhibitory loss of function mutants have been identified (R1021P) (Takeda et al., 2019), as well as a putative inhibitory phosphorylation site (T1024) (Perino et al., 2011). The resolution was sufficient to unambiguously build the three complementarity determining region (CDR) loops of NB7 that mediate target selectivity (Fig. 2E-F). The interface is primarily hydrophobic, with only 8 hydrogen bonds, and 1 electrostatic interaction among the 33 interfacial residues of NB7. A pocket formed between the helical domain and the ABD-RBD linker forms the majority of the interface, with extensive interactions with the long CDR3. The CDR1 loop packed up against the N-terminal section of the helical domain, with the CDR2 loop forming the interface with the regulatory motif. Previous study of oncogenic mutants in the regulatory motif of p110γ showed that increased dynamics mediated by these mutants increased kinase activity, putatively by breaking the autoinhibitory tryptophan lock in kα12 of the regulatory motif (Rathinaswamy et al., 2021c). Therefore, rigidifying the regulatory motif likely explains the molecular basis for how it prevents kinase activity. The nanobody interface is distinct from the predicted Gβγ interface (Rathinaswamy et al., 2023) and the experimentally resolved Ras interface (Pacold et al., 2000), explaining why it can still be membrane recruited by these stimuli.

The interface of NB7 with p110γ is distant from both the putative membrane binding surface, as well as the catalytic machinery of the kinase domain. To further understand how this nanobody could potently inhibit PI3K activity we examined any other potential modulators of PI3K activity localised in this region. There are two regulatory phosphorylation sites in the helical (Walser et al., 2013) and kinase domain (Perino et al., 2011) localised at the NB7 interface. This is intriguing as helical domain phosphorylation is activating, and kinase domain phosphorylation is inhibitory. This suggested a critical role in the regulation of p110γ is the dynamics of this kinase-helical interface. To fully define the role of NB7 in altering the dynamics of the helical domain we needed to study other modulators of helical domain dynamics.

p110γ activation by helical domain phosphorylation

To further understand the potential role of helical domain dynamics in controlling p110γ activity we examined the role of S582 helical domain phosphorylation by the protein kinase PKCβII (encoded by the gene PRKCB2, referred to as PKCβ for simplicity throughout the manuscript) (Walser et al., 2013). S582 is located on the interior of the helical domain and would not be expected to be exposed when the N-terminal region of the helical domain is folded (Fig. 3A). To understand this better at a molecular level, we purified a catalytic fragment of PKCβ and performed protein phosphorylation reactions on p110γ apo, p110γ-p84, and p110γ-p101. We identified a phosphorylated peptide containing S582, and surprisingly, we found an additional p110γ phosphorylation site at either S594/S595 (Fig. 3B, Fig. S4). The modification at this site results in a single phosphorylation event, but due to CID MS/MS fragmentation we cannot determine which site is modified, and we will refer it as S594/S595 throughout the manuscript. The S594/S595 site is also located in the N-terminal region of the helical domain, and is even more buried than S582, and would not be expected to be exposed when this region is folded (Fig. 3A). Dose response curves of PKCβ treatment was carried out for p110γ (Fig. 3C), p110γ-p84 (Fig. 3D), and p110γ-p101 (Fig. 3E). Both p110γ and p110γ-p84 showed similar dose response curves for PKCβ treatment, with similar curves for S582 and S594/S595. The p110γ-p101 complex was only very weakly phosphorylated, with <100-fold lower levels compared to p110γ and p110γ-p84 (Fig. 3E). This is consistent with the helical domain in p110γ being more rigid when bound to p101, compared to either bound to p84 or p110γ alone.

PKCβ leads to dual phosphorylation of internal sites in the helical domain, with selectivity for apo p110γ and p110γ-p84 over p110γ-p101

A. Putative phosphorylation sites mapped on the structure of p110γ (PDB: 7MEZ) and cartoon schematic. The regions are colored based on domain schematics featured in Fig 2A.

B. Raw MS spectra of the unphosphorylated and phosphorylated peptide for a region spanning 579-592 (RYESLKHPKAYPKL) and 593-607 (FSSVKWGQQEIVAKT). The putative phosphorylation sites in the sequence are shown in red, with the m/z theoretical (m/z t) and m/z experimental (m/z t) shown below each sequence.

C-E. Extracted traces and ratios of the intensity of extracted ion traces of different phosphorylation site peptides (Top to bottom: S594/S595 and S582) from (C) p110γ, (D) p110γ/p84 or (E) p110γ/p101 samples treated with increasing concentration of PKCβ according to the legend. The black traces in the ratio graphs are the intensity of the non-phosphorylated peptide, and the red traces in the ratio graphs are the intensity of the phosphorylated peptide.

To provide additional insight into the molecular mechanisms underlying p110γ phosphorylation we carried out hydrogen deuterium exchange mass spectrometry (HDX-MS) experiments on p110γ and phosphorylated p110γ (90.8% phosphorylated S594/595, 92% phosphorylated S582) (Fig. 4A). The full data underlying the experiment is available in the source data, and data processing information is in supplemental table 2. We have previously observed that the N-terminal region of the helical domain of apo p110γ (residues spanning 557-630aa) shows isotope profiles that are consistent with EX1 H/D exchange kinetics (Rathinaswamy et al., 2021b, 2021c; Vadas et al., 2013; Walser et al., 2013). This is indicative of cooperative unfolding of extended protein regions, with H/D exchange occurring faster than the refolding event. This region is where the PKCβ phosphorylation sites are located and may explain how the buried residues S582 and S594/S595 can be exposed to PKCβ. This is compatible with the observation that p110γ-p101 is protected from phosphorylation, as it does not show EX1 kinetics in this region, whereas both p110γ and p110γ-p84 do (Rathinaswamy et al., 2021a).

Activating phosphorylation at the helical domain leads to opening of the regulatory motif

A. HDX-MS comparing apo and phosphorylated p110γ. Significant differences in deuterium exchange are mapped on to the structure and cartoon of p110γ according to the legend (PDB: 7MEZ).

B. The graph of the #D difference in deuterium incorporation for p110γ, with each point representing a single peptide. Peptides colored in red are those that had a significant change in the mutants (greater than 0.4 Da and 5% difference at any timepoint, with a two tailed t-test p<0.01). Error bars are S.D. (n=3).

C. Representative bimodal distribution (EX1 kinetics) observed in the helical domain peptides of p110γ.

D. Representative p110γ peptides displaying increases in exchange in the phosphorylated state are shown. For all panels, error bars show SD (n[=[3)

E. Measurement of ATP to ADP conversion of phosphorylated and non-phosphorylated p110γ (1000 nM final concentration) ATPase activity in the absence (left) and presence of PIP2 membranes (5% phosphatidylinositol 4,5-bisphosphate (PIP2), 95% phosphatidylserine (PS)) activation (right). Significance is indicated by **(<0.001%), and ***(<0.0001%).

When we compared phosphorylated p110γ (>90.8% as measured by mass spectrometry at both sites) to unphosphorylated p110γ we observed extensive increases in dynamics in the C2, helical domain and kinase domain (Fig. 4A-D). The largest increases in exchange upon phosphorylation were located in the N-terminal region of the helical domain, with the peptides directly adjacent to the phosphorylation site showing almost complete deuterium incorporation at the earliest time points of exchange. This is indicative of significant disruption of the alpha helical secondary structure in this region. When we examined the exchange profiles in this region, they still underwent EX1 kinetics (Fig. 4C), however, phosphorylated p110γ was enriched in the more fully deuterated species. In addition to the regions in the helical domain, a portion of the regulatory motif of the kinase domain also showed increased deuterium exposure. This included the kα9-kα12 helices that surround the activation loop of p110γ. These increases in exchange were similar to those we had observed in a R1021C oncogenic activating mutant of PIK3CG (Rathinaswamy et al., 2021c).

To further explore the potential role of phosphorylation in mediating p110γ activity, we examined the kinase activity of p110γ under two conditions: basal ATP turnover, and with PIP2 containing lipid membranes. The experiments in the absence of PIP2 measure turnover of ATP into ADP and phosphate and is a readout of basal catalytic competency. Experiments with PIP2, measured ATP consumed in the generation of PIP3, as well as in non-productive ATP turnover. The p110γ enzyme in the absence of stimulators is very weakly active towards PIP2 substrate with only ∼2 fold increased ATP turnover compared to in the absence of membranes. This is consistent with very weak membrane recruitment of p110γ complexes in the absence of lipid activators (Rathinaswamy et al., 2023). PKCβ-mediated phosphorylation enhanced the ATPase activity of p110γ ∼2-fold in both the absence and presence of membranes (Fig. 4E). This suggests that the effect of phosphorylation is to change the intrinsic catalytic efficiency of phosphorylated p110γ, with limited effect on membrane binding.

Nanobody decreases p110γ phosphorylation

As NB7 bound at the interface of the helical and kinase domains that is exposed upon PKCβ phosphorylation of p110γ we hypothesized that the nanobody would likely alter phosphorylation. We carried out PKCβ phosphorylation of p110γ, p110γ-p101 and p110γ bound to NB7. The presence of NB7 showed a significant decrease in p110γ phosphorylation at both sites (Fig. 5A-B). We also wanted to determine whether p110γ phosphorylation reciprocally perturbed NB7 binding. BLI experiments showed that there was no detectable binding of NB7 to phosphorylated p110γ (Fig. 5C-D), consistent with phosphorylation disrupting the N-terminal region of the p110γ helical domain. In addition, lipid kinase assays using phosphorylated p110γ showed no detectable difference in activity when measured in the absence and presence nanobody (Fig. 5E), consistent with NB7 being unable to bind to phosphorylated p110γ.

Nanobody NB7 blocks PKCβ phosphorylation, and phosphorylation prevents nanobody binding

A. Extracted ion chromatograms for p110γ, p110γ-p101, and p110γ bound to NB7 are shown for the S594 or S595 phosphorylation site in p110γ. A bar graph showing the intensities of phosphorylated and non-phosphorylated p110γ peptide (593-607) for p110γ (black), p110γ with NB7 (red) and p110γp101 (purple) are shown to the right of the extracted ion chromatograms (n=3, right). In all experiments in panels A+B, PKCβ was present at 500 nM. Significance is indicated by ***(<0.0001%).

B. Extracted ion chromatograms for p110γ, p110γ-p101, and p110γ bound to NB7 are shown for the S582 phosphorylation site in p110γ. A bar graph showing the intensities of phosphorylated and non-phosphorylated p110γ peptide (579-592) p110γ (black), p110γ with NB7 (red) and p110γ-p101 (purple) are shown to the right of the extracted ion chromatograms (n=3, right). Significance is indicated by * (<0.01%), and ***(<0.0001%). The putative phosphorylation site is shown in red in the sequence above the bar graphs for both panel A+B.

C. Cartoon schematic of BLI analysis of the binding of immobilized His-NB7 to phosphorylated and non-phosphorylated p110γ.

D. Association curves for phosphorylated and non-phosphorylated p110γ (25nM) binding to His-NB7 are shown (n=3).

E. ATPase kinase activity assays comparing the activation/inhibition of phosphorylated and non-phosphorylated p110γ (1000 nM) with or without nanobody (3000 nM final) in the absence of PIP2 membranes. Significance is indicated by * (<0.05%), and NS (>0.05%).

Discussion

Here we find that the helical domain is a central regulator of the p110γ catalytic subunit of class IB PI3K, with modulation of helical dynamics through binding partners or PTMs able to either increase or decrease lipid kinase activity. These results expand on previous work defining the helical domain as a central regulator of class IA PI3Ks, where the nSH2 domain of the p85 regulatory subunits makes inhibitory interactions that significantly inhibit lipid kinase activity of all class IA catalytic subunits (p110α, p110β, and p110δ) (Mandelker et al., 2009; Miled et al., 2007; Burke and Williams, 2013; Burke, 2018). This inhibitory interaction in class IA PI3Ks is disrupted in human cancers (helical hotspot mutations in PIK3CA) (Samuels et al., 2004) and immune disorders (helical mutations in PIK3CD in APDS1) (Angulo et al., 2013; Lucas et al., 2014). Class IB PI3Ks are unique compared to class IA PI3Ks, as they are not inhibited by p101 and p84 regulatory subunits, but instead potentiate GPCR activation. This lack of inhibition is due to the distinct binding interface of class IB PI3K regulatory subunits compared to class IA regulatory subunits, with only class IA regulatory subunits making direct inhibitory interactions with the kinase and helical domains of p110 catalytic subunits (Rathinaswamy et al., 2021a). Here we show that a unique surface at the interface of the helical and kinase domains of p110γ is a potential site for the development of novel allosteric inhibitors that modulate p110γ activity.

The previously identified inhibitory nanobody (NB7) (Rathinaswamy et al., 2021b) bound with high affinity and inhibited all complexes of p110γ. The nanobody interface is distinct from how the nSH2 inhibits class IA PI3K activity, as its binding site is on the opposite face of the helical domain (Fig. 6A). The mechanism of inhibition is also distinct, as the nSH2-helical interaction plays a critical role in preventing membrane recruitment of inhibited class I PI3Ks, with removal of this interface either through pYXXM motif binding, or oncogenic mutations leading to increased membrane recruitment (Burke et al., 2012, 2011; Zhang et al., 2011). Analysis of the nanobody binding site compared to the structure of HRas-p110γ or the HDX-MS supported Alphafold-multimer prediction of Gβγ-p110γ (Rathinaswamy et al., 2023) shows that nanobody binding does not sterically block complex formation (Fig. 6B). This is consistent with it not blocking membrane recruitment by Ras/Gβγ.The nanobody inhibited ATP turnover both in solution and on membranes, suggesting that it prevents formation of a catalytically competent conformation of p110γ, but still allows for membrane recruitment. Further development of small molecule allosteric binding partners in this allosteric pocket between the kinase and helical domain may reveal the specific molecular interactions in this pocket that mediate inhibition.

Comparison of nanobody binding site compared to p85 inhibition of class IA PI3Ks and class IB activation sites

A. Comparison of the nanobody NB7 binding site in p110γ compared to the nSH2 inhibitory site in p110α (PDB: 3HHM) (Mandelker et al., 2009)

B. Comparison of the nanobody NB7 binding site in p110γ compared to the X-ray structure of the Ras binding site (PDB: 1HE8) (Pacold et al., 2000) and the Alphafold model of Gβγ bound to p110γ (Rathinaswamy et al., 2023).

C. Oncogenic mutations and post-translational modifications in spatial proximity to the nanobody binding site.

Oncogenic mutations are frequent in the class IA PI3Kα encoded by PIK3CA, with this being the 2nd most frequently mutated gene in human cancer (Lawrence et al., 2014). Mutations in p110γ encoded by PIK3CG in cancer are less frequent, however, they can still provide insight into regulatory mechanisms that control activity. Oncogenic mutations in the kinase domain (R1021C) and helical domain (E581K) are in close proximity to the nanobody binding site, and both would be expected to disrupt the stability of the helical domain or regulatory motif of the kinase domain (Fig. 6C). In addition to these mutations there are also multiple post-translational modifications that occur in this region, including inhibitory phosphorylation at T1024 (Perino et al., 2011), and activating phosphorylation at S582 (Walser et al., 2013). PKCβ is activated downstream of the IgE receptor in mast cells (Walser et al., 2013), but the full details of how this activates PI3K has been unclear. We identified an additional PKCβ phosphorylation site located in the helical domain (S594/S595) (Fig. 6C). Both the S582 and S594/S595 sites are not surface accessible and would require a transient opening of the helical domain for kinase accessibility. HDX-MS analysis of the helical domain of p110γ has shown that it is more dynamic than other class I PI3K isoforms (Burke and Williams, 2013; Walser et al., 2013), with the presence of the p101 regulatory subunit dramatically decreasing helical domain dynamics (Vadas et al., 2013). This putative mechanism of helical domain dynamics driving PKCβ phosphorylation is consistent with our observation that p101 subunits decreased p110γ phosphorylation >100-fold. PKCβ phosphorylation of p110γ leads to increased dynamics in both the helical and kinase domains with increased kinase activity, although only weakly compared to full activation by either membrane localised Ras or Gβγ. This increase was observed with both membrane and soluble substrate, so likely is not driven by altered membrane recruitment.

Overall, our biophysical and biochemical analysis of modulators of helical domain dynamics reveal the critical role of this domain in regulating class IB PI3K activity. This raises possibilities for development of small molecule modulators that may either increase or decrease helical domain dynamics, leading to either activation or inhibition. The high-resolution structure of an allosteric inhibitor nanobody provide initial insight into which pockets can specifically be targeted. Multiple ATP competitive p110γ selective inhibitors are in clinical trials for human cancers (Li et al., 2021), with many having significant side effects. The identification of novel inhibitory strategies provides new opportunities for targeting p110γ dysregulation in human disease.

Supporting information